Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-08 paper.Rmd
replace smooth benefit tables with figure; edits to paper

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
 - Richard K. Morgan^[World Bank, rick.morgan2@gmail.com]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
---

@@

Blair and Sambanis (hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.  Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.  Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.  We find that these claims are false.  Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.  The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative "non-theoretical" models that they compare to, and they use a random forest model that is appropriate for continuous, not discrete dependent variables..  In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.

@@

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.  Our belief is that these research decisions lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.  We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.

@@

This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. 
The reason to use a smoothed curve is that you have a continuous dependent variable (not the case here) which can produce any value of the dependent variable.  Wherein there is a discrete number of actual outcomes there is no justification for using smoothed ROC plots, nor for caculating statistics based on them.  Moreover, it is easy to compute the nonsmoothed ROC curve, and associated statistics, so computational ease does not justify this choice.

Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right.

@@

ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data. In this case, there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.  Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way. This ignores the fact that the actual ROC

Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves. The Average model outperforms the escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

table1_nosmooth %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of B\\&S Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
    label = "table1-nosmooth") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
```

As it turns out, this decision has a dramatic impact on the AUC-ROC calculations B&S use to support their empirical claims. As we show below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and this differentially impacts the models that are being compared. In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC. These results are shown in Tables 1 and 2.  The Goldstein, CAMEO, and Average model are everywhere superior to the theoretically informed escalation model.  Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.

To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper. This is not the case, and it is only mentioned in the sentence we quote above.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in Table 2. Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B&S Table 2 are actually not comparable toone another, and any conclusions drawn from such comparison are potentially incorrect.

We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for. The original B&S Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.

@@

# Results of the updated replication
We now turn to an examination of our replication that fixes the above issues.  The bottom line is that the changes we make turn on their heads the conclusions of B&S.

@@

table2_nosmooth %>%
  dplyr::select(-Model, -horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for escalation model with and without structural PITF contribution",
    label = "table2-fixed") %>%
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
  footnote("Differences from the original B\\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
```

@@

This is not the case anymore. It outperforms in the 1-month version and under-performs the escalation model in the 6-month version.

@@

Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model.

@@

```{r}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table1_benefit %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Smoothing advantage for B\\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
    label = "table1-benefit") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))

What impact did the ROC smoothing have overall on the results reported in B&S Tables 1 and 2? Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC. Positive values indicate that smoothing increased a model's apparent performance. The escalation model is the only model that consistently had a benefit from smoothing. For all eight other models, smoothing sometimes gave a benefit, sometimes not.

A more dramatic difference stands out when we consider the overall average impact of smoothing across all flavors of a model reported in the columns of the tables. The Average, Goldstein, Quad, PITF Split Population, and Weighted by PITF models are slightly hurt by smoothing, but on the order of less than 0.01 in absolute magnitude. The "With PITF Predictors" model is hurt on the order of 0.02, i.e. it appears substantially worse with smoothing. The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less. The escalation only model on the other hand has an average boost of 0.058 to its AUC-ROC from smoothing. Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.

The use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.

@@

# Additional concerns

@@

table2_full %>%
  select(horizon, column, N_test, cases) %>%
  pivot_wider(names_from = column, values_from = N_test) %>%
  arrange(horizon, desc(cases)) %>%
  rename(Cases = cases, Horizon = horizon) %>%
  select(Cases, Horizon, everything()) %>%
  arrange(desc(Cases), Horizon) %>%
  knitr::kable(
    "latex", booktabs = TRUE, digits = 0,
    caption = "Number of valid test predictions for each cell in B\\&S Table 2",
    label = "table2-N") %>%
  column_spec(3:7, width = "2cm") %>%
  collapse_rows(1:2, row_group_label_position = "stack")

@@

>>>>

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
 - Richard K. Morgan^[USAID, rick.morgan2@gmail.com. The views expressed are those of the author and do not reflect the official policy or position of USAID or the US Government.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
---

@@

Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.  Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.  Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.  We find that these claims are false.  Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.  The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure, the principal of which is that the performance results they report are based on smoothed performance curves, not the original unsmoothed curves. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented, and we also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.

@@

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation. These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts. We also note (e) additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.

Our belief is that these research decisions and issues lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.  We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.

@@

This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result appear step-like. Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right. The standard method is to compute the area under the curve AUC statistic on the original, non-smooth ROC curves that are shown on the right.

@@

ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values. The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.  Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way. This ignores the fact that the actual ROC **MIKE**

As we show below, the use of smoothed ROC curves to calculuate AUC has a dramatic impact on the results. To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper. As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in Table 2. The top portion of Table \ref{tab:table2-N} shows the number of valid test set predictions that can be scored for each model in the original B&S Table 2. For the 1-month data version, the number of predictions differs by up to 500 cases, and in the 6-month data version by around 100 cases. These numbers appear small enough to be not important, but crucially the already small number of positive cases also changes: the "With PITF Predictors" and "PITF Only" models lose 2 and 1 of 11 positive cases, respectively, in the 1- and 6-month data versions.

Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B&S Table 2 are actually not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.

table2_full %>%
  select(horizon, column, N_test, cases) %>%
  pivot_wider(names_from = column, values_from = N_test) %>%
  arrange(horizon, desc(cases)) %>%
  rename(Cases = cases, Horizon = horizon) %>%
  select(Cases, Horizon, everything()) %>%
  arrange(desc(Cases), Horizon) %>%
  knitr::kable(
    "latex", booktabs = TRUE, digits = 0,
    caption = "Number of valid test predictions for the escalation and structural alterantives models (B\\&S Table 2)",
    label = "table2-N") %>%
  column_spec(3:7, width = "2cm") %>%
  collapse_rows(1:2, row_group_label_position = "stack") 
```

We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for. The resulting number of cases for each model are shown in the bottom portion of the table. Since the sets of incomplete cases in the original version above do not completely overlap themselves, the common subset for all models is actually slightly smaller than the smallest N in the top portion of the table.

@@

Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.

# Results of the updated analysis

We now turn to an examination of our replication that fixes the above issues.  The bottom line is that the changes we make turn on their heads the conclusions of B&S.

## Is the Escalation model superior to the alternative ICEWS models?

Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves. Both the Average and CAMEO models outperform the Escalation model in almost all instances. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

table1_nosmooth %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Comparison of the escalation model to alternative ICEWS model using test set AUC-ROC, \\textit{without} smoothed ROC curves (Replication of B\\&S Table 1)",
    label = "table1-nosmooth") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
```

The key difference between our Table \ref{tab:table1-nosmooth} and the original B&S Table 1 is that we calculated AUC-ROC values for the original ROC curves, not smoothed ROC curves. B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are entirely conditional on the use of smoothed AUC-ROC. Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.

@@

table2_nosmooth %>%
  dplyr::select(-Model, -horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Comparison of escalation model to structural extensions, using test set AUC-ROC (Replication of B\\&S Table 2)",
    label = "table2-fixed") %>%
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
  footnote("Differences from the original B\\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
```

@@

The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguable performs better still.

@@

Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a maring in the 6-month data version.

@@

What impact did the ROC smoothing have overall on the results reported in Table \ref{tab:table1-nosmooth} (B&S Table 1) and Table \ref{tab:table2-fixed} (B&S Table 2)? Figure \ref{fig:benefit} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}. The red crosses mark the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.

```{r benefit, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table2_benefit <- read_csv("data/table2-smooth-benefit.csv")

benefit <- bind_rows(
  table1_benefit %>%
    rename(setting = Model) %>%
    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff"),
  table2_benefit %>%
    rename(setting = Model, Escalation = `Escalation Only`) %>%
    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff")
)

# mean benefit by model
means <- benefit %>%
  group_by(model) %>%
  summarize(avg_benefit = mean(diff))

ann_text <- tibble(horizon = "1 month", diff = 0.07, model = 8.6, label = "Mean")

model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
                  "PITF Only")
benefit %>%
  mutate(model = factor(model, levels = rev(model_levels))) %>%
  ggplot(aes(y = model)) + 
  #facet_wrap(~ horizon, scales = "free_x") +
  geom_point(aes(x = diff), alpha = 0.5, size = 2, color = "gray50") +
  geom_point(data = means, aes(x = avg_benefit), shape = 4, color = "red", size = 4) +
  labs(x = "AUC(smoothed ROC) - AUC(ROC)",
       y = "") +
  theme_bw() +
  geom_text(data = ann_text, aes(x = diff, label = label), col = "red") +
  geom_vline(xintercept = 0, linetype = 3) +
  theme(axis.text.y = element_text(size = 12))
```

In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.

@@

@@

@@

# Appendix

```{r smooth-benefit-tables}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table1_benefit %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Smoothing advantage for B\\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
    label = "table1-benefit") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
